---
title: "Activity Analysis"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This project will load and analyze some steps data from the activity.csv file.  
 
---
  
### Loading and preprocessing the data 

Let's begin our analysis by loading in the activity.csv file from inside the 
activity.zip file.

```{r loadcsv}
activity <- read.csv(unz("activity.zip", "activity.csv"))
```

---

### What is mean total number of steps taken per day?

Now we can calculate the total number of stes taken each day.

```{r totalsteps}
totalsteps <- aggregate(steps ~ date, activity, sum) 
```

Here we can now make our histogram for total steps per day, and calculate 
some statistics for the data set.

```{r plotandstats}
with(totalsteps, hist(x = steps, xlab = "Steps",
                      main = "Distribution of steps per day"))
avg <- mean(totalsteps$steps)
med <- median(totalsteps$steps)
cat(sprintf("Mean: %.2f\nMedian: %.2f", avg, med))
```

---

### What is the average daily activity pattern?

Now let's get the average steps per interval over all days, and plot them. We
also find the interval with the most steps on average, and print it.

```{r avgperday}
avgsteps <- aggregate(steps ~ interval, activity, mean) 
with(avgsteps, plot(interval, steps, type = "l", xlab = "5-minute Interval",
                    ylab = "Average Steps", main = "Mean steps per 5-minute interval"))
maxint <- with(avgsteps, subset(avgsteps, steps == max(steps)))$interval
```

The highest steps on average occur during interval `r maxint`, which
we can verify from the plot.

---

### Imputing missing values

Let's check on the NA values for the data set. First we should count the
number of NAs present.

```{r NAs}
NAs <- sum(is.na(activity$steps))
```

So we have `r NAs` NA values (`r round(100*(NAs / length(activity[[1]])), digits = 0)`%
of our data). Let's replace them with the means for that interval since we
have already calculated those values. Then we can plot and find the statistics
as we did before.

```{r missingtomean}
activitynew <- replace(activity, is.na(activity), values = avgsteps$steps)
totalstepsnew <- aggregate(steps ~ date, activitynew, sum) 
with(totalstepsnew, hist(x = steps, xlab = "Steps",
                         main = "Distribution of steps per day
                                (Imputed with mean)"))
avgnew <- mean(totalstepsnew$steps)
mednew <- median(totalstepsnew$steps)
cat(sprintf("Mean: %.2f\nMedian: %.2f", avgnew, mednew))
```

We see that the new imputed plot has a higher frequency near the average steps 
per day, which is expected since we added the average in place of all NA values.
We can also see the mean stayed the same (as expected) and the median became the
mean, which means we have more NA values than old median values, changing the median.

---

### Are there differences in activity patterns between weekdays and weekends?

Let's add a column to the activity dataset to indicate whether it is a weekday
or a weekend.

```{r weekfactor}
## We find all Saturday and Sunday, and set them equal to "weekend". Otherwise
##we make them equal to "weekday". We then convert to a factor and bind the col.
activitydays <- as.factor(ifelse(
    weekdays(strptime(activitynew$date, format = "%Y-%m-%d")) == "Saturday" |
    weekdays(strptime(activitynew$date, format = "%Y-%m-%d")) == "Sunday",
    "weekend", "weekday"))
activitynew <- cbind(activitynew, activitydays)
```

Now we can plot the weekdays and weekend days to compare the activity. We will use ggplot
to simplify the grouping.

```{r plotweekdays, results="hide"}
require(ggplot2)
avgstepsdays <- aggregate(steps ~ interval + activitydays, activitynew, mean)
qplot(interval, steps, data = avgstepsdays, col = activitydays, geom = "line")
```

We can see that on weekends, fewer steps are taken early as people will sleep in on 
weekends. There are also more steps during the day and later as people tend to 
move around rather than sit at work.